---
title: "Cost-Effectiveness of Continuous Glucose Monitoring vs. Blood Glucose Monitoring in Non-Insulin-Treated Type 2 Diabetes"
subtitle: "A Markov model with UKPDS-derived risk scaling, deterministic and probabilistic sensitivity analysis"
author: "Said Adjao"
date: today
format:
  html:
    toc: true
    number-sections: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup

library(tidyverse)

n_cycles <- 20
discount <- 0.03
states   <- c("Stable", "Complication", "Dead")
disc     <- 1 / (1 + discount)^(0:n_cycles)

md_hba1c     <- 0.42     # meta-analysis MD, magnitude
rrr_per_1pct <- 0.37     # UKPDS 35
rrr_scaled   <- 1 - (1 - rrr_per_1pct)^(md_hba1c / 1.0)

p_stable_comp_bgm <- 0.030
p_stable_comp_cgm <- 0.030 * (1 - rrr_scaled)
p_stable_death    <- 0.015
p_comp_death      <- 0.025

c_cgm <- 2500; c_comp <- 3000; c_stable <- 1500
u_stable <- 0.87; u_comp <- 0.75

make_tm <- function(p_event, p_sd, p_pd) {
  matrix(c(1 - p_event - p_sd, p_event, p_sd,
           0, 1 - p_pd, p_pd,
           0, 0, 1),
         nrow = 3, byrow = TRUE, dimnames = list(from = states, to = states))
}
run_markov <- function(tm, n_cycles) {
  tr <- matrix(0, n_cycles + 1, 3, dimnames = list(cycle = 0:n_cycles, state = states))
  tr[1, ] <- c(1000, 0, 0)
  for (t in 1:n_cycles) tr[t + 1, ] <- tr[t, ] %*% tm
  tr
}
tally <- function(trace, cost_vec, qaly_vec) {
  cc <- (trace %*% cost_vec) / 1000
  cq <- (trace %*% qaly_vec) / 1000
  c(cost = sum(cc * disc), qaly = sum(cq * disc))
}

tm_bgm <- make_tm(p_stable_comp_bgm, p_stable_death, p_comp_death)
tm_cgm <- make_tm(p_stable_comp_cgm, p_stable_death, p_comp_death)
qaly_vec <- c(u_stable, u_comp, 0)
cost_bgm <- c(c_stable, c_comp, 0)
cost_cgm <- c(c_stable + c_cgm, c_comp + c_cgm, 0)

res_bgm <- tally(run_markov(tm_bgm, n_cycles), cost_bgm, qaly_vec)
res_cgm <- tally(run_markov(tm_cgm, n_cycles), cost_cgm, qaly_vec)
d_cost_base <- res_cgm["cost"] - res_bgm["cost"]
d_qaly_base <- res_cgm["qaly"] - res_bgm["qaly"]
icer_base   <- d_cost_base / d_qaly_base
```

## Summary

This analysis evaluates the cost-effectiveness of continuous glucose monitoring
(CGM) versus blood glucose monitoring (BGM) in adults with type 2 diabetes not
treated with insulin. Unlike a direct clinical-event reduction, CGM's benefit
was modeled through a surrogate pathway: an independent meta-analysis estimated
CGM's effect on HbA1c (mean difference −0.42%), scaled to a microvascular
complication risk reduction using the dose-response relationship from UKPDS 35.

At a base-case annual cost of \$`r format(c_cgm, big.mark=",")`, the ICER was
**\$`r format(round(icer_base), big.mark=",")` per QALY** — exceeding the US
willingness-to-pay threshold and representing weaker value than a comparable
direct-acting therapeutic (SGLT2 inhibitors) modeled on the same framework.

## Background

CGM is a monitoring technology rather than a therapeutic agent: its value
depends on translating improved glycemic visibility into better long-term
outcomes via glucose control. This creates a structurally different economic
case than a drug with a directly measured clinical-event benefit, and this
analysis quantifies that difference explicitly.

## Methods

### From surrogate outcome to health states

CGM's treatment effect enters the model in two steps. First, an independent
meta-analysis of 8 RCTs estimated CGM's effect on HbA1c (MD −0.42%, 95% CI
−0.59 to −0.25). Second, this reduction was translated into a relative risk
reduction for microvascular complications using UKPDS 35, which found each 1%
absolute HbA1c reduction associated with a 37% (95% CI 33–41%) reduction in
microvascular complication risk, with no observed threshold. Scaling
exponentially to −0.42% yields a relative risk reduction of
**`r round(rrr_scaled * 100, 1)`%**, applied to the Stable→Complication
transition probability.

### Model structure

A three-state Markov model (Stable; Complication; Dead), `r n_cycles` annual
cycles, 3% discounting, mirrors the structure used for the companion SGLT2
inhibitor analysis.

```{r}
#| label: param-table

tibble(
  Parameter = c("HbA1c mean difference (CGM vs BGM)", "UKPDS RRR per 1% HbA1c",
                "Scaled RRR at −0.42%", "Baseline complication rate (annual)",
                "Background mortality (Stable)", "Mortality (Complication)",
                "CGM annual cost", "Complication annual cost",
                "Background care cost (Stable, annual)",
                "Utility: Stable", "Utility: Complication",
                "Discount rate", "Time horizon"),
  Value = c("−0.42% (95% CI −0.59, −0.25)", "37% (95% CI 33–41%)",
            paste0(round(rrr_scaled*100,1), "%"), "0.030", "0.015", "0.025",
            "$2,500", "$3,000", "$1,500",
            "0.87", "0.75", "3%", "20 years"),
  Source = c("Meta-analysis (8 RCTs)", "UKPDS 35 (Stratton et al., BMJ 2000)",
             "Calculated", "Illustrative", "Life tables", "Illustrative",
             "Device pricing", "Costing study", "Illustrative",
             "Utility study", "Utility study", "Convention", "—")
) %>% knitr::kable(caption = "Base-case model inputs.")
```

## Results

### Base case

```{r}
#| label: base-results

tibble(
  Arm = c("BGM", "CGM", "Incremental"),
  Cost = c(res_bgm["cost"], res_cgm["cost"], d_cost_base),
  QALYs = c(res_bgm["qaly"], res_cgm["qaly"], d_qaly_base)
) %>%
  mutate(Cost = paste0("$", format(round(Cost), big.mark=",")), QALYs = round(QALYs, 3)) %>%
  knitr::kable(caption = "Base-case costs and QALYs per patient (discounted).")
```

Base-case ICER: **\$`r format(round(icer_base), big.mark=",")`/QALY**.

### One-way sensitivity: UKPDS relationship

```{r}
#| label: owsa-ukpds

run_icer_ukpds <- function(rrr1) {
  rrr_s <- 1 - (1 - rrr1)^(md_hba1c / 1.0)
  tm_c  <- make_tm(p_stable_comp_bgm * (1 - rrr_s), p_stable_death, p_comp_death)
  rb <- tally(run_markov(tm_bgm, n_cycles), cost_bgm, qaly_vec)
  rt <- tally(run_markov(tm_c,   n_cycles), cost_cgm, qaly_vec)
  (rt["cost"] - rb["cost"]) / (rt["qaly"] - rb["qaly"])
}
tibble(Scenario = c("UKPDS low (33%)", "Base case (37%)", "UKPDS high (41%)"),
       ICER = sapply(c(0.33, 0.37, 0.41), run_icer_ukpds)) %>%
  mutate(ICER = paste0("$", format(round(ICER), big.mark=","))) %>%
  knitr::kable(caption = "ICER across the UKPDS 95% CI.")
```

The conclusion is unaffected across the UKPDS confidence interval.

### One-way sensitivity: CGM price

```{r}
#| label: owsa-price
#| fig-width: 8
#| fig-height: 4.5

run_icer_price <- function(price) {
  ct <- c(c_stable + price, c_comp + price, 0)
  rb <- tally(run_markov(tm_bgm, n_cycles), cost_bgm, qaly_vec)
  rt <- tally(run_markov(tm_cgm, n_cycles), ct, qaly_vec)
  (rt["cost"] - rb["cost"]) / (rt["qaly"] - rb["qaly"])
}
prices <- seq(100, 2500, by = 100)
price_df <- tibble(price = prices, icer = sapply(prices, run_icer_price))
threshold_price <- approx(price_df$icer, price_df$price, xout = 150000)$y

ggplot(price_df, aes(price, icer)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_hline(yintercept = 150000, linetype = "dashed", color = "red") +
  geom_vline(xintercept = threshold_price, linetype = "dotted") +
  scale_x_continuous(labels = scales::dollar) + scale_y_continuous(labels = scales::dollar) +
  labs(title = "ICER vs. Annual CGM Price", x = "Annual CGM cost", y = "ICER ($/QALY)") +
  theme_minimal()
```

CGM reaches \$150,000/QALY at an annual price of approximately
**\$`r format(round(threshold_price), big.mark=",")`** — a substantially larger
price reduction (from \$`r format(c_cgm, big.mark=",")`) than required for the
companion SGLT2 inhibitor model, reflecting the smaller absolute benefit
achievable through the surrogate pathway.

### Probabilistic sensitivity analysis

```{r}
#| label: psa
#| fig-width: 8
#| fig-height: 4.5

set.seed(123)
n_sims <- 5000
md_se    <- (0.59 - 0.25) / (2 * 1.96)
md_d     <- abs(rnorm(n_sims, 0.42, md_se))
ukpds_se <- (0.41 - 0.33) / (2 * 1.96)
ukpds_d  <- rnorm(n_sims, 0.37, ukpds_se)

beta_params  <- function(m, n) list(a = m*n, b = (1-m)*n)
gamma_params <- function(m, s) list(shape = (m/s)^2, rate = m/(s^2))
pc_d  <- with(beta_params(0.030, 200), rbeta(n_sims, a, b))
sd_d  <- with(beta_params(0.015, 200), rbeta(n_sims, a, b))
pd_d  <- with(beta_params(0.025, 200), rbeta(n_sims, a, b))
us_d  <- with(beta_params(0.87, 100),  rbeta(n_sims, a, b))
uc_d  <- with(beta_params(0.75, 100),  rbeta(n_sims, a, b))
ccg_d <- with(gamma_params(2500, 500), rgamma(n_sims, shape, rate))
ccm_d <- with(gamma_params(3000, 600), rgamma(n_sims, shape, rate))
cst_d <- with(gamma_params(1500, 300), rgamma(n_sims, shape, rate))

psa_cgm <- tibble(d_cost = numeric(n_sims), d_qaly = numeric(n_sims))
for (i in 1:n_sims) {
  rrr_i <- max(0, min(1, 1 - (1 - ukpds_d[i])^(md_d[i] / 1.0)))
  tm_b <- make_tm(pc_d[i], sd_d[i], pd_d[i])
  tm_c <- make_tm(pc_d[i]*(1 - rrr_i), sd_d[i], pd_d[i])
  qv <- c(us_d[i], uc_d[i], 0)
  cb <- c(cst_d[i], ccm_d[i], 0)
  cc <- c(cst_d[i]+ccg_d[i], ccm_d[i]+ccg_d[i], 0)
  rb <- tally(run_markov(tm_b, n_cycles), cb, qv)
  rt <- tally(run_markov(tm_c, n_cycles), cc, qv)
  psa_cgm$d_cost[i] <- rt["cost"] - rb["cost"]
  psa_cgm$d_qaly[i] <- rt["qaly"] - rb["qaly"]
}
p_150 <- round(100 * mean(psa_cgm$d_cost < 150000 * psa_cgm$d_qaly), 2)

ggplot(psa_cgm, aes(d_qaly, d_cost)) +
  geom_point(alpha = 0.12, color = "steelblue") +
  geom_abline(slope = 150000, intercept = 0, linetype = "dashed", color = "red") +
  scale_y_continuous(labels = scales::dollar) +
  labs(title = "Cost-Effectiveness Plane (5,000 simulations)",
       x = "Incremental QALYs", y = "Incremental Cost") + theme_minimal()
```

```{r}
#| label: ceac
#| fig-width: 8
#| fig-height: 4.5

thr <- seq(0, 500000, by = 10000)
ceac <- tibble(wtp = thr, p = sapply(thr, function(w) mean(psa_cgm$d_cost < w * psa_cgm$d_qaly)))
ggplot(ceac, aes(wtp, p)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_vline(xintercept = 150000, linetype = "dashed", color = "red") +
  scale_x_continuous(labels = scales::dollar) + scale_y_continuous(labels = scales::percent, limits=c(0,1)) +
  labs(title = "Cost-Effectiveness Acceptability Curve", x = "Willingness-to-pay ($/QALY)", y = "Probability cost-effective") +
  theme_minimal()
```

PSA probability of cost-effectiveness at \$150,000/QALY: **`r p_150`%**.

## Comparison to SGLT2 inhibitors

Both interventions were modeled on the same Markov framework, but their
mechanisms differ fundamentally: SGLT2 inhibitors reduce a directly measured
clinical event (HF hospitalization); CGM's benefit is mediated through a
surrogate marker requiring an additional risk-translation step. This is
reflected in the results — CGM's ICER, threshold price gap, and PSA probability
are all less favorable than SGLT2i's, despite both treatment effects being
statistically robust in their respective meta-analyses. This illustrates that
evidentiary certainty about effectiveness does not guarantee favorable
cost-effectiveness; the causal pathway's directness materially affects economic
value.

## Limitations

In addition to standard parameter uncertainty (tested via PSA), this model
carries a structural assumption not captured by sensitivity analysis: the
UKPDS 35 risk relationship derives from a 1977–1997 UK cohort, and its
applicability to a contemporary population may differ. The model assumes
linear, no-threshold scaling of risk reduction, consistent with UKPDS's own
reported findings, but this remains an extrapolation beyond directly observed
CGM trial data. As with the companion analysis, costs and utilities are
illustrative.

## Conclusion

CGM reduces HbA1c with a well-established effect, but under base-case
assumptions its cost-effectiveness for preventing microvascular complications
in non-insulin-treated T2DM is substantially weaker than a comparable
direct-acting therapy, driven by the indirectness of its benefit pathway
rather than uncertainty in the underlying evidence.


